Well-quasi-orderings on word languages

### 2024-08-11 src/infixes.tex
amalgamation?

In this section, we study the well-quasi-ordering of languages under the
\kl{infix relation}. As opposed to the \kl{prefix} and \kl{suffix} relations,
the \kl{infix relation} can yield to very complicated \kl{well-quasi-ordered}
languages. Formally, the following theorem shows that \emph{any} countable
quasi-ordering with finite initial segments can be embedded into the infix
relation of a language.

Let us recall that an \intro{order embedding} from a quasi-ordered set $(X,
\preceq)$ into a quasi-ordered set $(Y, \preceq')$ is a function $f \colon X
\to Y$ such that for all $x, y \in X$, $x \preceq y$ if and only if $f(x)
\preceq' f(y)$. When such an embedding exists, we say that $X$ \reintro{embeds
into} $Y$.

We say that a quasi-ordered set $(X, \preceq)$ is a \intro{partial ordering}
whenever the relation $\preceq$ is antisymmetric, that is $x \preceq y$ and $y
\preceq x$ implies $x = y$.

\begin{theorem}
    \label{infix-embedding:thm}
    Let $(X, \preceq)$ be a \kl{partially ordered} set,
    and $\Sigma \defined \set{a,b,c}$.
    Then the following are equivalent:
    \begin{enumerate}
        \item \label{infix-embedding-embeds:item} 
            $X$ \kl{embeds into} $(\Sigma^*, \infleq)$,
        \item \label{infix-embedding-count:item}
            $X$ is countable, and for every $x \in X$,
        $\setof{y \in X}{y \preceq x}$ is finite.
    \end{enumerate}
    Furthermore, if $X$ is recursively enumerable and $\preceq$ decidable,
    then
    the \kl{embedding} is computable.
\end{theorem}
\begin{proof}
    Let us first prove that \cref{infix-embedding-embeds:item} implies
    \cref{infix-embedding-count:item}. Let $f \colon X \to \Sigma^*$ be an
    \kl{embedding}. Note that $f$ is injective, because $X$ and $\Sigma^*$
    are \kl{partially
    ordered}. In particular, because $\dwset[\infleq]{w}$ is finite
    for every $w \in \Sigma^*$, then so is $\setof{y \in X}{y \preceq x}$
    for every $x \in X$. Similarly, $X$ must be countable
    because $\Sigma^*$ is.

@@

The reason why \kl{periodic words} built using a given period $x \in \Sigma^+$
are interesting for the \kl{infix relation} is that they naturally create
\kl{chains}. Indeed, if $x \in \Sigma^+$ is a finite word, then $\setof{x^p}{p
\in \Nat}$ is a \kl{chain} for the \kl{infix relation}. Note that in general,
the downwards closure of a chain is \emph{not} a chain. However, for the chains
generated using periodic words, the downwards closure
$\dwset[\infleq]{\setof{x^p}{p \in \Nat}}$ is a \emph{finite union} of
\kl{chains}. Because this set will appear in bigger equations,
we introduce the shorter notation 
$\intro*\InfPeriodChain{x}$ for the set of 
\kl{infixes} of words of the form $x^p$, where $p$ ranges over $\Nat$.

@@

\section{Infixes and Amalgamation systems}
\label{infixes-amalgamation:sec}

When proving \cref{infix-finite-automata:thm}, we have leveraged a powerful
combinatorial argument from \cref{pumping-periods:lem}, which made the proof
almost trivial. It turns out that there is a rather large family of systems for
which pumping arguments based on so-called \emph{minimal runs} exist: they are
called \emph{amalgamation systems}. Having done the heavy lifting on finite
automata, the rest of this section is mostly devoted to the introduction of
amalgamation systems and collecting the necessary pumping argument they enjoy.
Using this meta-proof, we will generalize \cref{infix-finite-automata:thm} to
context-free grammars, languages recognized by VASS, and more.

@@

Let us recall some examples of \kl{amalgamation systems}
arising in the literature.

\begin{example}
    Regular languages, context-free languages, and languages recognized by
    VASS are all recognized by amalgamation systems.
\end{example}

>>>>

In this section, we study the well-quasi-ordering of languages under the
\kl{infix relation}. As opposed to the \kl{prefix} and \kl{suffix} relations,
the \kl{infix relation} can yield to very complicated \kl{well-quasi-ordered}
languages. Formally, the upcoming \cref{infix-embedding:thm} shows that \emph{any} countable
quasi-ordering with finite initial segments can be embedded into the infix
relation of a language.

\AP
Let us recall that an \intro{order embedding} from a quasi-ordered set $(X,
\preceq)$ into a quasi-ordered set $(Y, \preceq')$ is a function $f \colon X
\to Y$ such that for all $x, y \in X$, $x \preceq y$ if and only if $f(x)
\preceq' f(y)$. When such an embedding exists, we say that $X$ \reintro{embeds
into} $Y$.
We say that a quasi-ordered set $(X, \preceq)$ is a \intro{partial ordering}
whenever the relation $\preceq$ is antisymmetric, that is $x \preceq y$ and $y
\preceq x$ implies $x = y$.

\begin{theorem}
    \label{infix-embedding:thm}
    Let $(X, \preceq)$ be a \kl{partially ordered} set,
    and $\Sigma \defined \set{a,b,c}$.
    Then the following are equivalent:
    \begin{enumerate}
        \item \label{infix-embedding-embeds:item} 
            $X$ \kl{embeds into} $(\Sigma^*, \infleq)$,
        \item \label{infix-embedding-count:item}
            $X$ is countable, and for every $x \in X$,
            its \kl{downwards closure}
            $\dwset[\preceq]{x}$ is finite.
    \end{enumerate}
\end{theorem}
\begin{proof}
    Let us first prove that \cref{infix-embedding-embeds:item} implies
    \cref{infix-embedding-count:item}. Let $f \colon X \to \Sigma^*$ be an
    \kl{embedding}. Note that $f$ is injective, because $X$ and $\Sigma^*$
    are \kl{partially
    ordered}. In particular, because $\dwset[\infleq]{w}$ is finite
    for every $w \in \Sigma^*$, then so is $\dwset[\preceq]{x}$
    for every $x \in X$. Similarly, $X$ must be countable
    because $\Sigma^*$ is.

@@

The reason why \kl{periodic words} built using a given period $x \in \Sigma^+$
are interesting for the \kl{infix relation} is that they naturally create
\kl{chains}. Indeed, if $x \in \Sigma^+$ is a finite word, then $\setof{x^p}{p
\in \Nat}$ is a \kl{chain} for the \kl{infix relation}. Note that in general,
the \kl{downwards closure} of a chain is \emph{not} a chain. However, for the chains
generated using periodic words, the \kl{downwards closure}
$\dwset[\infleq]{\setof{x^p}{p \in \Nat}}$ is a \emph{finite union} of
\kl{chains}. Because this set will appear in bigger equations,
we introduce the shorter notation 
$\intro*\InfPeriodChain{x}$ for the set of 
\kl{infixes} of words of the form $x^p$, where $p$ ranges over $\Nat$.

@@

\section{Infixes and Amalgamation Systems}
\label{infixes-amalgamation:sec}

When proving \cref{infix-finite-automata:thm}, we have leveraged a powerful
combinatorial argument from \cref{pumping-periods:lem}. It turns out that there
is a rather large family of systems for which pumping arguments based on
so-called \emph{minimal runs} exist: they are called \emph{amalgamation
systems} and were recently introduced by Anad, Schmitz, Sch\"{u}tze, and
Zetzsche \cite{ASZZ24}. Having done the heavy lifting on finite automata, the
rest of this section is mostly devoted to the introduction of \kl{amalgamation
systems} and collecting the necessary pumping argument they enjoy. Using this
meta-proof, we will generalize \cref{infix-finite-automata:thm} to context-free
grammars, languages recognized by vector addition systems with state (VASS),
and more. The goal of this section is not to introduce all of these systems, or
justify their usefulness, but to state and prove the following theorem.

@@

Let us recall some examples of \kl{amalgamation systems} given by the authors
of the original paper: regular languages \cite[Theorem 5.3]{ASZZ24}, VASS as a
consequence of \cite[Theorem 5.5]{ASZZ24}, context-free languages as a
consequence of \cite[Theorem 5.10]{ASZZ24}.

### 2024-08-11 src/prefixes.tex
amalgamation?

In this section, we study the well-quasi-ordering of languages under the prefix
relation. Let us immediately remark that the map $u \mapsto u^R$ that reverses
a word is an order-bijection between $(X^*, \prefleq)$ and $(X^*, \suffleq)$.
Therefore, we will focus on the prefix relation in the rest of this section.

Let us briefly restate the fact that some (even regular) languages are not
well-quasi-ordered by the prefix relation.

\begin{example}
    The set $L = \setof{a^nb}{n \in \Nat} \subseteq \set{a,b}^*$ is an infinite \kl{antichain} for the
    \kl{prefix relation}.
\end{example}

In order to characterize the existence of infinite antichains for the prefix
relation, we will introduce the following tree construction that
will be useful in the rest of this section.

@@

Because the alphabet $\Sigma$ is assumed to be finite, the \kl{tree of
prefixes} is finitely branching. Let us now observe how \kl{antichains} in the
prefix relation can be witnessed by infinite branches in the \kl{tree of
prefixes}.

@@

It is clear that an \kl{antichain branch} for a language $L$ yields an infinite
antichain, and the converse is quite easy to prove. Because the notion of
antichain branch is \kl{$\MSO$-definable} whenever $L$ is regular, we
immediately obtain decidability as a corollary of this lemma.

\begin{lemma}
    Let $L \subseteq \Sigma^*$ be a language. Then, $L$ contains an infinite
    \kl{antichain} if and only if there exists an \kl{antichain branch} for $L$.
\end{lemma}
\begin{proof}
    Assume that $L$ contains an \kl{antichain branch}. Let us construct an
    infinite \kl{antichain} as follows. We start with a set $A_0 \defined
    \emptyset$ and a node $v_0$ at the root of the tree. At step $i$, we
    consider a word $w_i$ such that $v_i$ is a \kl{prefix} of $w_i$, and $w_i
    \in L \setminus B$, which exists by definition of \kl{antichain branches}.
    We then set $A_{i+1} \defined A_i \cup \set{w_i}$. To compute $v_{i+1}$, we
    consider the largest prefix of $w_i$ that belongs to $B$, and set $v_{i+1}$
    to be the successor of this prefix in $B$. By an immediate induction, we
    conclude that for all $i \in \Nat$, $A_i$ is an \kl{antichain}, and that
    $v_i$ is a node in the \kl{antichain branch} $B$ such that $v_i$ is not a
    prefix of any word in $A_i$.

@@

\begin{corollary}
    \label{prefix-wqo-reg-decidable:cor}
    If $L$ is regular, then the existence of an infinite antichain is decidable.
\end{corollary}
\begin{proof}
    If $L$ is regular, then it is \kl{$\MSO$-definable}, and there 
    exists a formula $\varphi(x)$ in \kl{$\MSO$} that selects nodes 
    of the \kl{tree of prefixes} that belong to $L$. Now, to decide whether there
    exists an \kl{antichain branch} for $L$, we can simply check whether
    the following formula is satisfied:
    \begin{equation*}
        \exists B. 
        B \text{ is a branch } \land
        \forall x \in B, \exists y. y \text{ is a child of } x \land \varphi(y) \land y \not\in B
        \quad .
    \end{equation*}
    Because the above formula is an $\MSO$-formula over the infinite
    binary tree, whether it is satisfied is decidable.
    \textbf{TODO: cite}.
\end{proof}

@@

@@

    Conversely, assume that $L$ is \kl{well-quasi-ordered} by the \kl{prefix
    relation}. Let us define $S$ the set of words $w$ such that there exists
    two words $wu$ and $wv$ both in $L$ that are not comparable for the
    \kl{prefix relation}. Assume by contradiction that $S$ is infinite.
    Then, $S$ equipped with the \kl{prefix relation} is an infinite
    tree with finite branching, and therefore contains an infinite
    branch, which is by definition an \kl{antichain branch} for $L$.
    This contradicts the assumption that $L$ is \kl{well-quasi-ordered}.
    Now, consider $w$ be a maximal element for the \kl{prefix ordering}
    in $S$. By definition, all words in $L$ that contain $w$ as a prefix
    must be comparable for the \kl{prefix relation}. Therefore
    $\upset[\prefleq]{w} \cap L$ is a \kl{chain} for the \kl{prefix relation}.
    In particular, letting $S_{\max}$ be the set of all maximal elements
    of $S$,
    we conclude that 
    \begin{equation*}
        L \subseteq S \cup \bigcup_{w \in S_{\max}} \upset[\prefleq]{w} \cap L
        \quad .
    \end{equation*}
    Hence, that $L$ is finite union of \kl{chains}.
\end{proof}

>>>>

In this section, we study the well-quasi-ordering of languages under the
\kl{prefix relation}. Let us immediately remark that the map $u \mapsto u^R$
that reverses a word is an order-bijection between $(X^*, \prefleq)$ and $(X^*,
\suffleq)$, that is, $u \prefleq v$ if and only if $u^R \suffleq v^R$.
Therefore, we will focus on the prefix relation in the rest of this section, as
$(L, \prefleq)$ is \kl{well-quasi-ordered} if and only if $(L^R, \suffleq)$ is.

The next remark that we make is that $\Sigma^*$ is not \kl{well-quasi-ordered}
by the \kl{prefix relation} as soon as $\Sigma$ contains two distinct letters
$a$ and $b$. As an example of infinite \kl{antichain}, we can consider the set
of words $a^n b$ for $n \in \Nat$. As mentioned in the introduction, there are
some languages however that are \kl{well-quasi-ordered} by the \kl{prefix
relation}. A simple example being the (regular) language $a^* \subseteq
\set{a,b}^*$, which is order-isomorphic to natural numbers with their usual
orderings $(\Nat, \leq)$.

In order to characterize the existence of infinite \kl{antichains} for the
\kl{prefix relation}, we will introduce the following tree construction that
will be useful in the rest of this section.

@@

We will use this \kl{tree of prefixes} to find simple witnesses
of the existence of infinite \kl{antichains} in the \kl{prefix relation}
for a given language $L$, namely by introducing \kl{antichain branches}.

@@

Let us illustrate the notion of \kl{antichain branch} over the alphabet $\Sigma
= \set{a,b}$, and the language $L = a^* b$. In this case, the set $a^*$ (which
is a branch of the \kl{tree of prefixes}) is an \kl{antichain branch} for $L$.
This holds because for any $a^k$ in the branch, $a \prefleq a^kb$ which is in
$L$ and not in the branch. In general, the existence of an \kl{antichain
branch} for a language $L$ implies that $L$ contains an infinite
\kl{antichain}, and because the alphabet $\Sigma$ is assumed to be finite, one
can leverage the fact that the \kl{tree of prefixes} is finitely branching to
prove that the converse holds as well.

\begin{lemma}
    \label{antichain-branches-prefix:lem}
    Let $L \subseteq \Sigma^*$ be a language. Then, $L$ contains an infinite
    \kl{antichain} if and only if there exists an \kl{antichain branch} for $L$.
\end{lemma}
\begin{proof}
    Assume that $L$ contains an \kl{antichain branch}. Let us construct an
    infinite \kl{antichain} as follows. We start with a set $A_0 \defined
    \emptyset$ and a node $v_0$ at the root of the tree. At step $i$, we
    consider a word $w_i$ such that $v_i$ is a \kl{prefix} of $w_i$, and $w_i
    \in L \setminus B$, which exists by definition of \kl{antichain branches}.
    We then set $A_{i+1} \defined A_i \cup \set{w_i}$. To compute $v_{i+1}$, we
    consider the largest prefix of $w_i$ that belongs to $B$, and set $v_{i+1}$
    to be the successor of this prefix in $B$. By an immediate induction, we
    conclude that for all $i \in \Nat$, $A_i$ is an \kl{antichain}, and that
    $v_i$ is a node in the \kl{antichain branch} $B$ such that $v_i$ is not a
    prefix of any word in $A_i$.

@@

One immediate application of \cref{antichain-branches-prefix:lem} is that
\kl{antichain branches} can be described inside the \kl{tree of prefixes} by a
monadic second order formula (\kl{$\MSO$-formula}), allowing us to leverage the
decidability of $\MSO$ over infinite binary trees \cite[Theorem 1.1]{RAB69}.

\begin{corollary}
    \label{prefix-wqo-reg-decidable:cor}
    If $L$ is regular, then the existence of an infinite antichain is decidable.
\end{corollary}
\begin{proof}
    If $L$ is regular, then it is \kl{$\MSO$-definable}, and there 
    exists a formula $\varphi(x)$ in \kl{$\MSO$} that selects nodes 
    of the \kl{tree of prefixes} that belong to $L$. Now, to decide whether there
    exists an \kl{antichain branch} for $L$, we can simply check whether
    the following formula is satisfied:
    \begin{equation*}
        \exists B. 
        B \text{ is a branch } \land
        \forall x \in B, \exists y. y \text{ is a child of } x \land \varphi(y) \land y \not\in B
        \quad .
    \end{equation*}
    Because the above formula is an $\MSO$-formula over the infinite
    $\Sigma$-branching tree, whether it is satisfied is decidable
    as an easy consequence of the decidability of $\MSO$ over infinite binary
    trees
    \cite[Theorem 1.1]{RAB69}.
\end{proof}

@@

For the proof of the following theorem, we will use special notations to
describe the \intro{upwards closure} of a set $S$ for a relation $\preceq$,
which is defined as $\upset[\preceq]{S} \defined \setof{y \in \Sigma^*}{
\exists x \in S. x \preceq y}$. Anticipating the use of the symmetric notion of
\intro{downwards closure}, let us introduce its notation 
straight away as follows: $\dwset[\preceq]{S} \defined \setof{y
\in \Sigma^*}{ \exists x \in S. y \preceq x}$. Abusing notations, we will
write $\upset{w}$ and $\dwset{w}$ for the upwards and downwards closure of a
single element $w$, and omit the ordering relation when it is clear from the
context.

@@

    Conversely, assume that $L$ is \kl{well-quasi-ordered} by the \kl{prefix
    relation}. Let us define $S$ the set of words $w$ such that there exists
    two words $wu$ and $wv$ both in $L$ that are not comparable for the
    \kl{prefix relation}. Assume by contradiction that $S$ is infinite.
    Then, $S$ equipped with the \kl{prefix relation} is an infinite
    tree with finite branching, and therefore contains an infinite
    branch, which is by definition an \kl{antichain branch} for $L$.
    This contradicts the assumption that $L$ is \kl{well-quasi-ordered}.
    Now, consider $w$ be a maximal element for the \kl{prefix ordering}
    in $S$. By definition, all words in $L$ that contain $w$ as a prefix
    must be comparable for the \kl{prefix relation}. Therefore,
    $(\upset[\prefleq]{w}) \cap L$ is a \kl{chain} for the \kl{prefix relation}.
    In particular, letting $S_{\max}$ be the set of all maximal elements
    of $S$,
    we conclude that 
    \begin{equation*}
        L \subseteq S \cup \bigcup_{w \in S_{\max}} (\upset[\prefleq]{w}) \cap L
        \quad .
    \end{equation*}
    Hence, that $L$ is finite union of \kl{chains}.
\end{proof}
